---
title: "Merkel_ChIP-seq_MohsenRastgooShahrestani"
author: "Mohsen Rastgoo Shahrestani"
date: "2024-11-21"
output: html_document
---

1. Load the following libraries:
```{r}
library(GenomicFeatures)
library(GenomicRanges)
library(org.Hs.eg.db)
library(ggplot2)
library(ChIPseeker)

library(TxDb.Hsapiens.UCSC.hg19.knownGene)
txdb <- TxDb.Hsapiens.UCSC.hg19.knownGene

library(clusterProfiler)
library(ReactomePA)
```

2) Access the information for hg19 ChIP-seq dataset of Merkel cell polyomavirus
transcription factors in GEO database ( https://www.ncbi.nlm.nih.gov/geo/ ) using
getGEOInfo.
```{r}
hg19 <- getGEOInfo(genome = "hg19", simplify = T)
```

3) you have to look for the GSM ids from series GSE69877, and keep the ones for
“ST-HA_ChIP-seq”, “EP400_ChIP-seq rep1” and “Max_ChIP-seq rep1”. Download
them using downloadGSMbedFiles.
Those were;
ST-HA_ChIP-seq <- GSM1711856
EP400_ChIP-seq rep1 <- GSM1711857
Max_ChIP-seq rep1 <- GSM1711858

```{r}
target_gsms <- c("GSM1711856", "GSM1711857", "GSM1711858")
```

```{r}
downloadGSMbedFiles(GSM = target_gsms, destDir = getwd())
```

4) Create a list called files with the paths of the data from the Merkel cell polyomavirus
dataset

```{r}
myfiles <- list(STHA = "D:/Molecular Biology/3rdSemester/Integrative Biology/Practicals/GSM1711856_STHA_1.peaks_peaks.bed.gz",
                P400 = "D:/Molecular Biology/3rdSemester/Integrative Biology/Practicals/GSM1711857_p400_1.peaks_peaks.bed.gz",
                MAX = "D:/Molecular Biology/3rdSemester/Integrative Biology/Practicals/GSM1711858_MAX_1.peaks_peaks.bed.gz")
print(myfiles)
```

```{r}
STHApeak <- readPeakFile(myfiles[[1]])
p400peak <- readPeakFile(myfiles[[2]])
Maxpeak <- readPeakFile(myfiles[[3]])
```

1. ChIP profiling
Look at the data for each single TF:
● Look at the GRanges objects. How many peaks are there? Which TF has the highest
number of peaks?
```{r}
num_peaks_STHA <- length(STHApeak)
num_peaks_p400 <- length(p400peak)
num_peaks_MAX <- length(Maxpeak)
peak_counts <- c(STHA = num_peaks_STHA, p400 = num_peaks_p400, MAX = num_peaks_MAX)
most_peaks <- names(which.max(peak_counts))
cat("TF with the highest number of peaks:", most_peaks, "\n")
```

● How long are peaks on average? Explore the GRanges slots (you can access slots
with “@”).

```{r}
lengths_STHA <- STHApeak@ranges@width
lengths_p400 <- p400peak@ranges@width
lengths_MAX <- Maxpeak@ranges@width

mean_length_STHA <- mean(lengths_STHA)
mean_length_p400 <- mean(lengths_p400)
mean_length_MAX <- mean(lengths_MAX)
print(c(mean_length_STHA, mean_length_p400, mean_length_MAX))
```

ChIP peaks coverage plot:
● How are peaks located over the whole genome? Provide a plot to visualize this
information.
```{r}
covplot(STHApeak, weightCol = "V5", title = "STHApeak Coverage")
covplot(p400peak, weightCol = "V5", title = "p400peak Coverage")
covplot(Maxpeak, weightCol = "V5", title = "Maxpeak Coverage")
```

Profile of ChIP peaks binding to TSS/different regions:
● Detect the ChIP peaks binding to TSS regions. Generate the tagMatrix objects
(maybe, it is better to keep TFs together and create a tagMatrix list, you may use
lapply to do it quickly)
```{r}
subset_peaks <- list(
  STHA = STHApeak[sample(1:length(STHApeak), size = 1000)],
  p400 = p400peak[sample(1:length(p400peak), size = 1000)],
  MAX = Maxpeak[sample(1:length(Maxpeak), size = 1000)]
)

tagMatrix_list <- lapply(subset_peaks, function(peaks) {
  getTagMatrix(peaks, TxDb = txdb, by = "gene", type = "start_site", upstream = 3000, downstream = 3000)
})
```


● Detect the same ChIP peaks (i.e. generate another tagMatrix list) using binning
method (use parameter nbin).
```{r}
binned_tagMatrix_list <- lapply(subset_peaks, function(peaks) {
  getTagMatrix(peaks, TxDb = txdb, by = "gene", type = "start_site", upstream = 3000, downstream = 3000, nbin = 800)
})

```

● Detect the ChIP peaks (i.e. generate another tagMatrix list) binding to a different
region of your choice.
```{r}
tagMatrix_body_list <- lapply(subset_peaks, function(peaks) {
  getTagMatrix(peaks, TxDb = txdb, by = "gene", type = "body", nbin = 50)
})
```

● Plot the heatmap of the tagMatrix for TSS regions. Is it possible to plot more heatmaps
at the same time?
I used a subset for this question since it took a lot of time for the original data and I could not manage to do it with my laptop.
```{r}

lapply(names(tagMatrix_list), function(name) {
  tagHeatmap(tagMatrix_list[[name]], title = paste(name, "TSS Binding Heatmap (Subset)"))
})
```

● Plot the average profile of the tagMatrix for binning method.
```{r}
lapply(names(binned_tagMatrix_list), function(name) {
  plotAvgProf(
    binned_tagMatrix_list[[name]],
    xlim = c(1,800),
    title = paste(name, "Average Profile (Binned)")
  )
})
```

2. Peak Annotation

● Annotate peaks. Also here, keep objects together inside a list as before.
```{r}
peak_annotations <- lapply(list(STHA = STHApeak, p400 = p400peak, MAX = Maxpeak), function(peaks) {
  annotatePeak(peaks, TxDb = txdb, tssRegion = c(-3000, 3000), verbose = FALSE)
})
```

Visualize Genomic Annotation:
● Explore the annotation information using the dedicated plots.
Pie and Bar plot;
```{r}
plotAnnoPie(peak_annotations[[1]])
plotAnnoPie(peak_annotations[[2]])
plotAnnoPie(peak_annotations[[3]])
```

```{r}
plotAnnoBar(peak_annotations[[1]])
plotAnnoBar(peak_annotations[[2]])
plotAnnoBar(peak_annotations[[3]])
```

3. ChIP peak data set comparison

Functional profiles comparison:
● Collect the annotated genes and keep them in a list by datasets.
```{r}
genes = lapply(peak_annotations, function(i) as.data.frame(i)$geneId)
names(genes) = sub("_", "\n", names(genes))
```

● Annotated genes can be analyzed by enrichment analyses. Compare the KEGG
pathways across the ChIP-Seq experiments (use function compareCluster).
● Provide the plot useful to visualize the enrichment analysis results.
```{r}
compKEGG <- compareCluster(geneCluster   = genes,
                         fun           = "enrichKEGG",
                         pvalueCutoff  = 0.05,
                         pAdjustMethod = "BH")
dotplot(compKEGG, showCategory = 15, title = "KEGG Pathway Enrichment Analysis")
```

Overlap of annotated genes:
● Provide a plot to visualize the overlaps between the annotated genes.
```{r}
vennplot(genes)
```

4. Statistical testing of ChIP seq overlap

Peak overlap enrichment analysis:
● Detect the peak overlap enrichment using as queryPeak “MAX_1” versus the others.
```{r}
queryPeak <- peak_annotations[["MAX"]]  
targetPeaks <- peak_annotations[c("STHA", "p400")]


enrichPeakOverlap(
  queryPeak = queryPeak@anno,  
  targetPeak = lapply(targetPeaks, function(x) x@anno),  
  TxDb = txdb,
  pAdjustMethod = "BH",
  nShuffle = 50,
  chainFile = NULL,
  verbose = FALSE
)
```

5. Define the cell signaling pathways impacted by ST protein of the virus

Look at the genes annotated in all the three TFs and perform an enrichment analysis on the
Reactome enrichment pathways:
● Intersect the annotated genes from the three TFs (STHA, EP400 e MAX).
```{r}
genes <- lapply(peak_annotations, function(i) {
  unique(as.data.frame(i)$geneId)
})
common_genes <- Reduce(intersect, genes)  
cat("Number of common genes:", length(common_genes), "\n")
```

● Use enrichPathway to perform the enrichment analysis on the Reactome pathway
database.
```{r}
library(ReactomePA)
reactome_results <- enrichPathway(
  gene = common_genes,
  organism = "human",  
  pvalueCutoff = 0.05,
  readable = TRUE
)

head(as.data.frame(reactome_results))
```

● Visualize the dotplot of the results.
```{r}
library(clusterProfiler)
dotplot(reactome_results, showCategory = 10, title = "Reactome Pathway Enrichment")
```



